---
title: "FEMA Hazard Mitigation projects analysis"
author: "Andrew Ba Tran"
output:
  html_document:
    code_folding: hide
---

The analysis done by The Post for the story [The Ring in Ashes](https://www.washingtonpost.com/nation/2021/11/15/fema-hazard-mitigation-funding/) can be reproduced using FEMA's [Hazard Mitigation Assistance Projects]((https://www.fema.gov/openfema-data-page/hazard-mitigation-assistance-projects-v2)) open data and joined with with FEMA's [Disaster Declarations Summaries]((https://www.fema.gov/openfema-data-page/disaster-declarations-summaries-v2)) data to determine which designated disasters occurred between three time periods 


```{r setup, warning=F, message=F}
library(tidyverse)
library(tidycensus)
library(lubridate)
library(DT)
library(knitr)
library(viridis)
library(scales)

knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(fig.width = 8)
knitr::opts_chunk$set(fig.height = 3)

# load up cleaned up data
mega_df <- readRDS("../../data/clean/mega_df_pt2.RDS") %>%
  mutate(declaration_year=year(declarationDate)) %>% 
  filter(programArea=="HMGP")


```

Of the $11 billion FEMA has allocated for this program over the past decade, only $1.6 billion has been spent.

```{r one}
# 1. What percent of money is unspent
share1 <- mega_df %>%
   #filter(programFy>2010) %>%
   mutate(year_declared=year(declarationDate)) %>%
   filter(year_declared>2010) %>% 
   filter(programArea=="HMGP") %>%
   group_by(status) %>%
   summarize(amount=sum(projectAmount, na.rm=T)) %>%
   mutate(percent_money=round(amount/sum(amount, na.rm=T)*100,1))

kable(share1, format.args = list(big.mark = ","))
```


How long counties wait on average to complete their FEMA-funded resilience projects.

Projects that closed in 2020 took an average of 7 years to get funded and completed, one year longer than it took a decade ago. 

This is counting from a declared disaster dates to dates projects closed by year of closure.

```{r disasters_closed_annually_chart}
mega_df %>%
  mutate(year_closed=year(dateClosed)) %>% 
  mutate(disaster_closed_interval=interval(declarationDate, dateClosed)) %>% 
  mutate(months_disaster_closed=disaster_closed_interval %/% months(1)) %>% 
  mutate(months_disaster_closed=case_when(
    months_disaster_closed < 0 ~ 0,
    TRUE ~ months_disaster_closed
  )) %>% 
  filter(year_closed>2010 & year_closed < 2021) %>% 
  group_by(year_closed) %>%
  #filter(initially_approved_year>2010) %>% 
  summarize(total=n(),
            months_disaster_closed=round(mean(months_disaster_closed, na.rm=T),1)) %>% 
  mutate(years=months_disaster_closed/12) %>% 
  filter(year_closed>2010) %>% 
  ggplot(aes(x=year_closed, y=years)) +
  
  geom_bar(position="stack", stat="identity") +
  #facet_wrap(~still_open) +
  scale_y_continuous(labels = scales::comma) +
  labs(title="Years it took for projects to close",
       subtitle="By project close year",
       caption="Data: FEMA",
       y="Years",
       x="") +
  theme_minimal()
```

It takes nearly two years for projects in rural counties to get approved.

```{r rural_approval}
# import county census data (poverty, urban, race)
county_data <- read_csv("../../data/clean/county_combined.csv") %>% 
  mutate(pov_quantile= ntile(pctpov, 4))  %>% 
  mutate(pop_quantile=ntile(poverty_population, 4))

# import county names
county_names <- read_csv("../../data/clean/county_names.csv")

county_data <- county_data %>% left_join(county_names)

mega_df %>%
     mutate(fipsStateCode=case_when(
         nchar(stateNumberCode)==1 ~ paste0("0", stateNumberCode),
         TRUE ~ as.character(stateNumberCode)
     )) %>%
     mutate(fipsCountyCode=case_when(
         nchar(countyCode)==1 ~ paste0("00", countyCode),
         nchar(countyCode)==2 ~ paste0("0", countyCode),
         is.na(countyCode) ~ "000",
         TRUE ~ as.character(countyCode)
     )) %>%
     mutate(GEOID=paste0(fipsStateCode, fipsCountyCode)) %>%
     left_join(county_data, by="GEOID") %>% 
     mutate(ur=case_when(
         (urban_rural > 0 & urban_rural <= 2) ~ "1. Urban",
         (urban_rural > 2 & urban_rural <= 4) ~ "2. Suburban",
         (urban_rural > 4 & urban_rural <= 6) ~ "3. Rural",
         TRUE ~ "4. Unclassified"
     )) %>%
     mutate(initially_approved_year=year(approval_date)) %>%
     mutate(initially_declared_approved_interval=interval(declarationDate, approval_date)) %>% 
     mutate(months_declared_approved=initially_declared_approved_interval %/% months(1)) %>% 
     filter(initially_approved_year>2010) %>% 
     group_by(ur) %>%
     summarize(total=n(),
               costs=sum(projectAmount, na.rm=T),
               months_declared_approved=round(mean(months_declared_approved, na.rm=T),1)) %>% 
  filter(ur=="3. Rural") %>% 
  kable()
```

How long it takes counties to close out projects by poverty percentile.

```{r poor}
# poor communities
mega_df %>%
     mutate(fipsStateCode=case_when(
         nchar(stateNumberCode)==1 ~ paste0("0", stateNumberCode),
         TRUE ~ as.character(stateNumberCode)
     )) %>%
     mutate(fipsCountyCode=case_when(
         nchar(countyCode)==1 ~ paste0("00", countyCode),
         nchar(countyCode)==2 ~ paste0("0", countyCode),
         is.na(countyCode) ~ "000",
         TRUE ~ as.character(countyCode)
     )) %>%
     mutate(GEOID=paste0(fipsStateCode, fipsCountyCode)) %>%
    mutate(year_closed=year(dateClosed)) %>% 
    #mutate(disaster_closed_interval=interval(declarationDate, approval_date)) %>% 
    mutate(disaster_closed_interval=interval(declarationDate, dateClosed)) %>% 
    mutate(months_disaster_closed=disaster_closed_interval %/% months(1)) %>% 
    left_join(county_data, by="GEOID") %>%
    #filter(year(approval_date)>2010) %>%    
    filter(year_closed>2010) %>% 
     group_by(pov_quantile) %>%
     summarize(total=n(),
               costs=sum(projectAmount, na.rm=T),
               months_declared_closed=round(mean(months_disaster_closed, na.rm=T),1))  %>% 
  filter(!is.na(pov_quantile)) %>% 
  kable()
```



How long it takes to close out projects in counties where majority of the population is either white or non-white.

```{r nonwhite}
# white or non-white
mega_df %>%
     mutate(fipsStateCode=case_when(
         nchar(stateNumberCode)==1 ~ paste0("0", stateNumberCode),
         TRUE ~ as.character(stateNumberCode)
     )) %>%
     mutate(fipsCountyCode=case_when(
         nchar(countyCode)==1 ~ paste0("00", countyCode),
         nchar(countyCode)==2 ~ paste0("0", countyCode),
         is.na(countyCode) ~ "000",
         TRUE ~ as.character(countyCode)
     )) %>%
    mutate(GEOID=paste0(fipsStateCode, fipsCountyCode)) %>%
    mutate(year_closed=year(dateClosed)) %>% 
    mutate(disaster_closed_interval=interval(declarationDate, approval_date)) %>% 
    #mutate(disaster_closed_interval=interval(declarationDate, dateClosed)) %>% 
    mutate(months_disaster_closed=disaster_closed_interval %/% months(1)) %>% 
    select(-majority) %>% 
    left_join(county_data, by="GEOID") %>%
    filter(year(approval_date)>2010) %>% 
#    filter(year_closed>2010) %>% 
    group_by(majority) %>%
    summarize(total=n(),
               costs=sum(projectAmount, na.rm=T),
               months_declared_closed=round(mean(months_disaster_closed, na.rm=T),1)) %>% 
  filter(!is.na(majority)) %>% 
  kable()
```

Counties with at least one project that took longer than 10 years to close out where majority of the population is either white or non-white.

```{r plusten_race}
tenplus <- mega_df %>%
     mutate(fipsStateCode=case_when(
         nchar(stateNumberCode)==1 ~ paste0("0", stateNumberCode),
         TRUE ~ as.character(stateNumberCode)
     )) %>%
     mutate(fipsCountyCode=case_when(
         nchar(countyCode)==1 ~ paste0("00", countyCode),
         nchar(countyCode)==2 ~ paste0("0", countyCode),
         is.na(countyCode) ~ "000",
         TRUE ~ as.character(countyCode)
     )) %>%
    mutate(GEOID=paste0(fipsStateCode, fipsCountyCode)) %>%
    mutate(year_closed=year(dateClosed)) %>% 
    #mutate(disaster_closed_interval=interval(declarationDate, dateClosed)) %>% 
    mutate(disaster_closed_interval=interval(declarationDate, approval_date)) %>% 
    mutate(months_disaster_closed=disaster_closed_interval %/% months(1)) %>% 
    select(-majority) %>% 
    left_join(county_data, by="GEOID") %>%
    filter(months_disaster_closed>120) %>% 
  mutate(year_approved=year(approval_date)) %>% 
  filter(year_approved>2010)
  #filter(year_closed>2010)

# Counties with at least one project that took longer than 10 years to close out
tenplus %>% 
    count(GEOID) %>% 
    mutate(project=1) %>% 
    right_join(county_data, by="GEOID") %>% 
    group_by(majority) %>%
    summarize(total=n(),
              counties=sum(project, na.rm=T)) %>% 
  ungroup() %>% 
  mutate(percent_of_counties=round(counties/total*100,1)) %>% 
  kable()
```

Counties with projects that took longer than 10 years to close out in counties by poverty quantiles.

```{r plusten_pov}
tenplus <- mega_df %>%
     mutate(fipsStateCode=case_when(
         nchar(stateNumberCode)==1 ~ paste0("0", stateNumberCode),
         TRUE ~ as.character(stateNumberCode)
     )) %>%
     mutate(fipsCountyCode=case_when(
         nchar(countyCode)==1 ~ paste0("00", countyCode),
         nchar(countyCode)==2 ~ paste0("0", countyCode),
         is.na(countyCode) ~ "000",
         TRUE ~ as.character(countyCode)
     )) %>%
    mutate(GEOID=paste0(fipsStateCode, fipsCountyCode)) %>%
    mutate(year_closed=year(dateClosed)) %>% 
    mutate(disaster_closed_interval=interval(declarationDate, dateClosed)) %>% 
    mutate(months_disaster_closed=disaster_closed_interval %/% months(1)) %>% 
    select(-majority) %>% 
    left_join(county_data, by="GEOID") %>%
    filter(months_disaster_closed>120) %>% 
    filter(year_closed>2010)

tenplus %>% 
    mutate(project=1) %>% 
    group_by(GEOID) %>% 
    summarize(projects=sum(project, na.rm=T)) %>%   
    right_join(county_data, by="GEOID") %>% 
    group_by(pov_quantile) %>%
    summarize(total=n(),
              counties=sum(projects, na.rm=T)) %>% 
  filter(!is.na(pov_quantile)) %>% 
  kable()
```

Average number of disasters affecting counties while waiting for relevant hazard mitigation projects.

```{r disaster_overlap}
project_types <- read_csv("../../data/raw/project_types.csv")

mega_df <- mega_df %>%
  left_join(project_types)

mega_df <- mega_df %>%
  mutate(incident_count=str_count(disaster_wide_type, incidentType)) 

mega_df %>%
  filter(subappType!="Management Costs") %>%
  filter(subappType!="Planning") %>%
  filter(programFy>2010) %>%
  group_by(incidentType) %>%
  summarize(average_disaster_count=round(mean(incident_count, na.rm=T),1)) %>%
  arrange(-average_disaster_count) %>% 
  head(3) %>% 
  kable()


#Hurricane
hurricanes <- 
     mega_df %>%
     filter(subappType!="Management Costs") %>%
     filter(subappType!="Planning") %>%
     filter(programFy>2010) %>% filter(incidentType=="Hurricane")

stc <- mega_df %>% 
  filter(year(approval_date)>2010) %>% 
  filter(county=="St. Charles")
#Fire

#Severe Ice Storm
ice <- 
     mega_df %>%
     filter(subappType!="Management Costs") %>%
     filter(subappType!="Planning") %>%
     filter(programFy>2010) %>% filter(incidentType=="Severe Ice Storm")
# too many are still open

#Flood
Flood <- 
     mega_df %>%
     filter(subappType!="Management Costs") %>%
     filter(subappType!="Planning") %>%
     filter(programFy>2010) %>% filter(incidentType=="Flood")


#Tornado
tornados <- 
     mega_df %>%
     filter(subappType!="Management Costs") %>%
     filter(subappType!="Planning") %>%
     filter(programFy>2010) %>% filter(incidentType=="Tornado")

typh<- 
     mega_df %>%
     filter(subappType!="Management Costs") %>%
     filter(subappType!="Planning") %>%
     filter(programFy>2010) %>% filter(incidentType=="Typhoon")

```

Disasters compared to projects by county type.


```{r yup, fig.height=6, fig.width=8}

disasters <- read_csv("../../data/raw/DisasterDeclarationsSummaries090321.csv") %>% 
  mutate(GEOID=paste0(fipsStateCode, fipsCountyCode)) 

# import county census data (poverty, urban, race)
county_data <- read_csv("../../data/clean/county_combined.csv") %>% 
  mutate(pov_quantile= ntile(pctpov, 4))  %>% 
  mutate(pop_quantile= ntile(poverty_population,4))
# import county names (whoops)
county_names <- read_csv("../../data/clean/county_names.csv")

county_data <- county_data %>% left_join(county_names)


county_disaster_list <- disasters %>% 
  ungroup() %>% 
  select(-placeCode, -designatedArea, -hash, -id) %>% 
  unique() %>% 
  mutate(declaration_year=year(declarationDate)) %>% 
  filter(declaration_year>2010)

write_csv(county_disaster_list, "../../data/clean/county_disaster_list_dupe.csv", na="")

county_disaster_list <- read_csv("../../data/clean/county_disaster_list_dupe.csv") %>% 
  unique()

county_disaster_count <- county_disaster_list %>% 
  count(GEOID, name="disasters") %>% 
  full_join(county_data) %>% 
  mutate(ur=case_when(
    (urban_rural > 0 & urban_rural <= 2) ~ "1. Urban",
    (urban_rural > 2 & urban_rural <= 4) ~ "2. Suburban",
    (urban_rural > 4 & urban_rural <= 6) ~ "3. Rural",
    TRUE ~ "4. Unclassified"
  )) %>% 
  filter(!is.na(NAME))

county_projects_count <- mega_df %>% 
  filter(programArea=="HMGP") %>%
  filter(year(declarationDate)>2010) %>% 
#  filter(is.na(dateClosed) | year(dateClosed)>2010) %>% 
  mutate(fipsStateCode=case_when(
    nchar(stateNumberCode)==1 ~ paste0("0", stateNumberCode),
    TRUE ~ as.character(stateNumberCode)
  )) %>%
  mutate(fipsCountyCode=case_when(
    nchar(countyCode)==1 ~ paste0("00", countyCode),
    nchar(countyCode)==2 ~ paste0("0", countyCode),
    is.na(countyCode) ~ "000",
    TRUE ~ as.character(countyCode)
  )) %>%
  mutate(GEOID=paste0(fipsStateCode, fipsCountyCode)) %>% 
  count(GEOID, name="projects")

county_projects_disasters <- full_join(county_disaster_count, county_projects_count) %>% 
  filter(!is.na(NAME))

ggplot(county_projects_disasters, aes(x=disasters, y=projects, color=ur)) +
  geom_point()+
  #geom_smooth(method='lm', formula= y~x, se=FALSE) +
  geom_abline(intercept=0, slope=1) +
  geom_smooth(method=lm) +
  labs(title="Number of projects versus disasters by county type") +
  theme_minimal()
```

Urban counties are twice as likely on average to receive funding after a disaster than rural areas.

```{r ratio}
county_projects_disasters %>% 
  mutate(projects2=case_when(
    is.na(projects) ~ 0,
    TRUE ~ as.numeric(projects)
  ),
  disasters2=case_when(
    is.na(disasters) ~ 0,
    TRUE ~ as.numeric(disasters)
  )) %>% 
  mutate(ratio=projects2/disasters2) %>% 
  group_by(ur) %>% 
  summarize(average_ratio = mean(ratio, na.rm=T)) %>% 
  kable()
```

Half of all rural counties in the U.S. experienced a disaster over the past 10 years, yet didn't have a single mitigation project.

```{r no_money}


disasters_counties <- read_csv("../../data/raw/DisasterDeclarationsSummaries090321.csv") %>% 
  mutate(GEOID=paste0(fipsStateCode, fipsCountyCode)) %>% 
  filter(incidentType=="Biological") %>% 
  filter(year(declarationDate)>2010) %>% 
  select(GEOID) %>% 
  unique()


counties_only <- mega_df %>% 
  filter(year(declarationDate)>2010) %>% 
  select(GEOID) %>% 
  unique() %>% 
  left_join(county_data, by="GEOID") %>% 
  mutate(ur=case_when(
      (urban_rural > 0 & urban_rural <= 2) ~ "1. Urban",
      (urban_rural > 2 & urban_rural <= 4) ~ "2. Suburban",
      (urban_rural > 4 & urban_rural <= 6) ~ "3. Rural",
      TRUE ~ "4. Unclassified"
    )) %>%
    count(ur, name="counties_with_projects")

no_mitigation <- readRDS("../../data/clean/no_mitigation.RDS")
county_data_none <- county_data %>%
  mutate(pov_quantile= ntile(pctpov, 4)) %>%
  filter(GEOID %in% no_mitigation)

total_ur <- county_data %>% 
  count(urban_rural, name="total_counties")
# 
# county_data_none %>%
#   count(urban_rural) %>% 
#   left_join(total_ur) %>% 
#   mutate(percent=round(n/total_counties*100,1)) %>% 
#   rename(counties_with_no_projects=n,
#          counties=total_counties) %>% 
#   filter(!is.na(urban_rural)) %>% 
#   kable()

county_data_none <- county_data %>%
  mutate(pov_quantile= ntile(pctpov, 4)) %>%
  filter(GEOID %in% no_mitigation)

total_ur <- county_data %>% 
  count(urban_rural, name="total_counties")

county_data_none %>%
  count(urban_rural) %>% 
  left_join(total_ur) %>% 
mutate(ur=case_when(
    (urban_rural > 0 & urban_rural <= 2) ~ "1. Urban",
    (urban_rural > 2 & urban_rural <= 4) ~ "2. Suburban",
    (urban_rural > 4 & urban_rural <= 6) ~ "3. Rural",
    TRUE ~ "4. Unclassified"
  )) %>%
  group_by(ur) %>% 
  summarize(total_counties=sum(total_counties),
            n=sum(n)) %>% 
  left_join(counties_only) %>% 
  filter(ur!="4. Unclassified") %>% 
  mutate(percent=round(n/total_counties*100,1)) %>% 
  rename(type=ur,
         `total counties`=total_counties,
         `qualified counties with no projects`=n,
         `counties with projects`=counties_with_projects,
         `percent counties with no projects`=percent) %>% 
  kable()

```

Southern states wait longer.

```{r states}
states <- read_csv("../../data/raw/states.csv")

mega_df %>%
  mutate(fipsStateCode=case_when(
    nchar(stateNumberCode)==1 ~ paste0("0", stateNumberCode),
    TRUE ~ as.character(stateNumberCode)
  )) %>%
  mutate(fipsCountyCode=case_when(
    nchar(countyCode)==1 ~ paste0("00", countyCode),
    nchar(countyCode)==2 ~ paste0("0", countyCode),
    is.na(countyCode) ~ "000",
    TRUE ~ as.character(countyCode)
  )) %>%
  mutate(GEOID=paste0(fipsStateCode, fipsCountyCode)) %>%
  mutate(year_closed=year(dateClosed)) %>% 
  mutate(disaster_closed_interval=interval(declarationDate, dateClosed)) %>% 
  mutate(months_disaster_closed=disaster_closed_interval %/% months(1)) %>% 
  filter(year(dateClosed)>2010) %>% 
  left_join(states, by=c("fipsStateCode"="st_fips")) %>% 
  group_by(region.y) %>% 
  summarize(total=n(),
            costs=sum(projectAmount, na.rm=T),
            months_declared_closed=round(mean(months_disaster_closed, na.rm=T),1)) %>% 
  arrange(-months_declared_closed) %>% 
  filter(!is.na(region.y)) %>% 
  kable()
```

Specifically West South Central, which consists of Arkansas, Louisiana, Oklahoma, and Texas.

```{r states_division}

mega_df %>%
  mutate(fipsStateCode=case_when(
    nchar(stateNumberCode)==1 ~ paste0("0", stateNumberCode),
    TRUE ~ as.character(stateNumberCode)
  )) %>%
  mutate(fipsCountyCode=case_when(
    nchar(countyCode)==1 ~ paste0("00", countyCode),
    nchar(countyCode)==2 ~ paste0("0", countyCode),
    is.na(countyCode) ~ "000",
    TRUE ~ as.character(countyCode)
  )) %>%
  mutate(GEOID=paste0(fipsStateCode, fipsCountyCode)) %>%
  mutate(year_closed=year(dateClosed)) %>% 
  mutate(disaster_closed_interval=interval(declarationDate, dateClosed)) %>% 
  mutate(months_disaster_closed=disaster_closed_interval %/% months(1)) %>% 
  filter(year(dateClosed)>2010) %>% 
  left_join(states, by=c("fipsStateCode"="st_fips")) %>% 
  group_by(division) %>% 
  summarize(total=n(),
            costs=sum(projectAmount, na.rm=T),
            months_declared_closed=round(mean(months_disaster_closed, na.rm=T),1)) %>% 
  arrange(-months_declared_closed) %>% 
  filter(!is.na(division)) %>% 
  kable()
```
